---
title: "Representing gene/UMI counts per spot"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: | 
  Here, we read in the SM-omics manuscript data and for the 'brain' data,
  extract the gene/umi per spot and plot
output:
  pdf_document:
  html_notebook: default
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
---


```{r}
library(tidyverse) # Load tidyverse pacakges
```

data files needed to run are available at SCP: https://singlecell.broadinstitute.org/single_cell/study/SCP979/
please download:
10005CN48_C1_downsamp.tsv
10005CN48_D1_downsamp.tsv
10005CN48_E1_downsamp.tsv
10015CN60_E2_downsamp.tsv
10015CN84_C2_downsamp.tsv
10015CN84_D2_downsamp.tsv
10005CN48_C1_stdata_under_tissue_IDs.txt
10005CN48_D1_stdata_under_tissue_IDs.txt
10005CN48_E1_stdata_under_tissue_IDs.txt
10015CN60_E2_stdata_under_tissue_IDs.txt
10015CN84_C2_stdata_under_tissue_IDs.txt
10015CN84_D2_stdata_under_tissue_IDs.txt
```{r}
path <- '../smomics_data/' # set path to results directory
```

Define sample information.
```{r}
# Sample-condition mapper
cond_sm <- list('10005CN48_C1'='ST',
                '10005CN48_D1'='ST',
                '10005CN48_E1'='ST',
                '10015CN84_D2'='SM-Omics',
                '10015CN84_C2'='SM-Omics',
                '10015CN60_E2'='SM-Omics')

# Sample:seq-depth mapper
seq_depth <- list('10015CN84_D2'=23113376,
                  '10015CN60_E2'=33876251,
                  '10015CN84_C2'=28174220,
                  '10005CN48_C1'=71022006,
                  '10005CN48_D1'=75643539,
                  '10005CN48_E1'=58631758)
```

Read in tsv counts files.
```{r}
samples_list <- c('10005CN48_C1', '10005CN48_D1', '10005CN48_E1', '10015CN84_D2', '10015CN84_C2', '10015CN60_E2')

files_dict <- samples_list %>% lapply(function(filename){
  read.table(file.path(path, paste0(filename, '_downsamp_stdata.tsv.gz')), row.names=1) %>% t()
})

names(files_dict) <- samples_list
```


Identify spots 'inside' tissues. Seems this can be done using the '_stdata_under_tissue_IDs.txt.gz' files and
taking the first row as spots inside tissue.
```{r}
under_tissues <- samples_list %>% lapply(function(filename){
  
  con <- file(file.path(path, paste0(filename, '_stdata_under_tissue_IDs.txt.gz')),"r")
  
  #
  if(filename=='10005CN48_E1'){ # this file requires specific handling
    first_line <- readLines(con,n=1) %>% strsplit('\t') # get first line and split by tab delimiter
    
    
    spots <- first_line[[1]] %>% lapply(function(spot){
      # for each element, split into two by '_' delimiter
      spot <- strsplit(spot, '_')[[1]]
      
      # round first and second values
      spot1 <- round(as.numeric(spot[[1]]))
      spot2 <- round(as.numeric(spot[[2]]))
      
      # recombine with 'x' delimiter
      sprintf('%sx%s', spot1, spot2)
    }) %>% unlist()
  
  } else{
    # get first line, replace '_' with 'x' and split by tab delimiter
    first_line <- readLines(con,n=1) %>% gsub(pattern = '_', replacement = 'x') %>% strsplit('\t')
    spots <- first_line[[1]]

  }
  
  close(con)
  
  return(spots)

})


names(under_tissues) <- samples_list

# check number of spots in tissues and total number of spots.
under_tissues %>% sapply(length)
files_dict %>% sapply(ncol)

```

Define a function to extract features (genes/umis) per spot.

Note that the thresholding is slightly different to the sm-omics jupyter notebooks:

- Column thresholding is > seq-depth/1E6. This is slightly more precise than the notebooks which
use approximately this value
- Row thresholding value (100) is applied to all spots rather than separately inside and outside 
```{r}
get_features_per_spot <- function(feature='gene', threshold=500){
  
  samples_list %>% lapply(function(sample){
    all_counts = files_dict[[sample]]
  
    # Remove spots (columns) with too small total count
    threshold <- seq_depth[[sample]]/1E6
    counts_above_thresh <- all_counts[,colSums(all_counts)>threshold]
  
    spots_inside <- under_tissues[[sample]]
    
    # remove genes (rows) with too small total count
    grouped_features <- counts_above_thresh[rowSums(counts_above_thresh)>threshold,] %>%
      data.frame() %>%
      tibble::rownames_to_column('gene_id') %>% # move rowname to column to preserve
      pivot_longer(cols=-gene_id, names_to='spot', values_to='count') %>% # pivot into longer form
      filter(count>0) %>% # remove rows with zero count
      mutate(spot=gsub('^X', '', spot)) %>% # rename spot to remove leading 'X' from data.frame column conversion
      mutate(inside=ifelse(spot %in% spots_inside, 'Inside tissue', 'Outside tissue')) %>% # add 'inside' column
      group_by(spot, inside)
    
    if(feature=='gene'){ # just tally instances
      out <- grouped_features %>%
        tally()
    } else if(feature=='umi'){ # sum counts
      out <- grouped_features %>%
        summarise(n=sum(count))
    } else{
      stop('feature must be gene or umi')
    }
    
    out %>% mutate(sample_name=sample) # add sample name info
    
  }) %>%
    bind_rows() %>%
    mutate(condition=recode(sample_name, !!!cond_sm)) # recode sample name to make condition column
}

genes_per_spot <- get_features_per_spot('gene')
umis_per_spot <-  get_features_per_spot('umi')

```

Function to plot results
```{r}

plot_feature_per_spot <- function(feature_per_spot){
  feature_per_spot %>%
    ggplot(aes(condition, n, group=sample_name)) +
    geom_violin(position=position_dodge(width=1)) +
    geom_boxplot(width=0.1, outlier.shape=NA, position=position_dodge(width=1)) +
    facet_wrap(~inside) +
    theme_bw(base_size=20) +
    theme(panel.grid=element_blank()) +
    xlab('')
}


p <- plot_feature_per_spot(genes_per_spot) + ylab('Genes per spot')
print(p)
ggsave('./gene_per_spot_dist.png')

p <- plot_feature_per_spot(genes_per_spot) + scale_y_log10() + ylab('UMIs per spot')
print(p)
ggsave('./umis_per_spot_dist.png')

  

```

Plot aggregated (mean) genes per spot for comparison.
```{r}
p <- genes_per_spot %>%
  bind_rows() %>%
  group_by(sample_name, inside) %>%
  summarise(mean_n=mean(n)) %>%
  mutate(condition=recode(sample_name, !!!cond_sm)) %>%
  ggplot(aes(inside, mean_n, colour=condition)) +
  geom_point(position=position_dodge(width=0.5)) +
  theme_bw(base_size=15) +
  theme(aspect.ratio=1.5, 
        panel.grid=element_blank()) +
  scale_color_discrete(name='') +
  xlab('') +
  ylab('Mean genes per spot')

print(p)
ggsave('./gene_per_spot_mean.png')



  
```
